---
title: "Selection Bias and Type I Error Inflation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Selection Bias and Type I Error Inflation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(dosopt)
```

## Background

In a two-stage dose optimization design, Stage 1 selects the dose with the higher observed utility score. Because selection is correlated with observed response rates, the response rate on the selected dose is **upward biased**. This bias inflates the Type I error rate in the confirmatory Stage 2 test.

## Analytical Bias Formula

For binary efficacy, the Stage 1 selection bias is

$$\text{Bias}_1 = \frac{\text{Cov}(X, U)}{\sigma_U \sqrt{n_1}} \cdot \frac{1}{\sqrt{\pi}} \exp\!\left(-\frac{\lambda_u^2 n_1}{4\sigma_U^2}\right),$$

and the combined Stage 1 + Stage 2 bias is attenuated by $n_1 / (n_1 + n_2)$.

```{r bias-binary}
# First get sample size (n per arm in Stage 1)
ss <- calc_sample_size_utility_approx(
  pL = 0.30, qL = 0.70, delta = 0.10, d = 0.15,
  phi = 0, alpha_L = 0.80, alpha_H = 0.80
)

# Bias calculation (Stage 1 = n per arm; Stage 2 = 2x Stage 1)
bias_res <- calc_bias(
  pL    = 0.30, qL = 0.70,
  delta = 0.10, d  = 0.15,
  phi   = 0,
  n1    = ss$n,
  n2    = 2 * ss$n,
  lambda_u = ss$lambda_u
)
print(bias_res)
```

### Bauer Bound

An upper bound on bias that does not require knowing $\text{Cov}(X, U)$:

$$|\text{Bias}_1| \leq \sqrt{\frac{p(1-p)}{n_1 \pi}}$$

This bound is always reported alongside the analytical estimate.

## Type I Error Inflation

Given bias, the actual Type I error for a one-sample Z-test at nominal level $\alpha$ is

$$\text{TIE} = 1 - \Phi\!\left(z_{1-\alpha} - \frac{\text{Bias}_{\text{comb}}}{\text{SE}_0}\right).$$

```{r t1e-z}
t1e_z <- calc_type1_error(
  bias_stage1 = bias_res$bias_stage1,
  n1 = ss$n, n2 = 2 * ss$n,
  p0 = 0.30, alpha = 0.05, test = "z"
)
print(t1e_z)
```

For an exact binomial test, a corrected critical value is used:

```{r t1e-binom}
t1e_b <- calc_type1_error(
  bias_stage1 = bias_res$bias_stage1,
  n1 = ss$n, n2 = 2 * ss$n,
  p0 = 0.30, alpha = 0.05, test = "binomial"
)
print(t1e_b)
```

## Bias as a Function of Stage 1 Sample Size

```{r bias-vs-n}
n1_seq <- seq(30, 150, by = 10)
bias_seq <- sapply(n1_seq, function(n1) {
  calc_bias(0.30, 0.70, 0.10, 0.15, phi = 0,
            n1 = n1, n2 = 120, lambda_u = ss$lambda_u)$bias_stage1
})

# Quick summary
cat("n1 range:", range(n1_seq), "\n")
cat("Bias range:", round(range(bias_seq), 4), "\n")
cat("Bias at n1=60:", round(bias_seq[n1_seq == 60], 4), "\n")
```

Bias decreases as $n_1$ increases, because larger stage 1 samples make the selection event ($\bar{U}_H - \bar{U}_L > \lambda_u$) increasingly deterministic.

## Landmark Survival Type I Error

For a landmark survival analysis (e.g., 12-month PFS rate):

```{r t1e-landmark}
t1e_lm <- calc_type1_error_landmark(
  bias_stage1 = bias_res$bias_stage1,
  n1 = ss$n, n2 = 2 * ss$n,
  p0 = 0.55,   # landmark survival rate under H0
  rho = 0.6,   # efficacy-survival surrogacy correlation
  alpha = 0.05
)
print(t1e_lm)
```

## Two-Arm Cox Regression Type I Error

For a confirmatory randomized trial using Cox regression, bias propagates through a delta-method chain: $\text{Bias}(\hat{p}) \to \text{Bias}(\hat{\lambda}) \to \text{Bias}(\log \widehat{HR}) \to \text{Bias}(Z_{\text{Cox}})$.

```{r t1e-cox}
t1e_cox <- calc_type1_error_cox(
  bias_stage1 = bias_res$bias_stage1,
  n1 = ss$n, n2 = 2 * ss$n,
  p0 = 0.30, rho = 0.6,
  lambda0 = log(2) / 12,   # median survival 12 months under H0
  n_confirm = 300,
  alpha = 0.05
)
print(t1e_cox)
```

## Summary

The selection-induced bias is modest for typical Phase 2 sample sizes but can inflate Type I error by 1–3 percentage points in confirmatory trials. Investigators should either:

1. Pre-specify an adjusted alpha for the confirmatory trial, or
2. Inflate Stage 1 sample size to reduce bias below an acceptable threshold.

Use `calc_bias()` and `calc_type1_error*()` together to explore these trade-offs during protocol development.
