---
title: "Introduction to dosopt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to dosopt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(dosopt)
```

## Motivation

FDA's **Project Optimus** initiative requires sponsors to characterize the dose–response relationship for both efficacy and safety, not merely identify the maximum tolerated dose. The **dosopt** package implements a frequentist, one-stage randomized dose optimization design that selects the optimal dose using a composite **utility score** balancing binary efficacy and safety endpoints.

## The Utility Score

For a patient with binary efficacy $X$ and binary toxicity $Y$, the utility score is

$$U = u_1 X(1-Y) + u_2 XY + u_3 (1-X)(1-Y) + u_4 (1-X)Y,$$

where $u_1 \geq u_2, u_3 \geq u_4$ and the independence constraint $u_1 - u_2 - u_3 + u_4 = 0$ holds.

Canonical weights derived from margins $(\delta, d)$ are

$$u_2 = \frac{1}{1+r},\quad u_3 = \frac{r}{1+r}, \quad r = \frac{\delta}{d}.$$

```{r utility-scores}
# Utility weights for delta = 0.10, d = 0.15  =>  r = 2/3
u <- calc_utility(r = 0.10 / 0.15)
print(u)
```

## Joint Outcome Probabilities

Given marginal rates $(p, q)$ and Pearson correlation $\phi$ between $X$ and $Y$:

```{r calc-pi}
pi_vec <- calc_pi(p = 0.35, q = 0.65, phi = 0)
names(pi_vec) <- c("XY=11", "XY=10", "XY=01", "XY=00")
print(round(pi_vec, 4))

# Feasible correlation range
phi_bounds(p = 0.35, q = 0.65)
```

## Sample Size: Margin-Based (Mode 1)

Provide lower-dose rates, efficacy margin $\delta$, and toxicity margin $d$:

```{r mode1}
res <- calc_sample_size_utility_approx(
  pL    = 0.30,
  qL    = 0.70,
  delta = 0.10,
  d     = 0.15,
  phi   = 0,
  alpha_L = 0.80,
  alpha_H = 0.80
)
print(res)
```

The output includes sample size per arm `n`, selection threshold `lambda_u`, and probability of correct selection (PCS) for each dose.

## Sample Size: Partial-Direct (Mode 2)

If you have partial information about the higher dose:

```{r mode2}
res2 <- calc_sample_size_utility_approx(
  pL    = 0.30, qL = 0.70,
  pH_L  = 0.20, qH_L = 0.05,   # higher-dose rates if lower is optimal
  pL_H  = 0.40, qL_H = 0.15,   # lower-dose rates if higher is optimal
  phi   = 0,
  alpha_L = 0.80, alpha_H = 0.80,
  mode  = 2
)
print(res2)
```

## Interactive Shiny App

For an interactive GUI covering all design modes:

```r
run_dosopt_app()
```

## Next Steps

- **Exact methods**: `vignette("sample-size-methods")`
- **Bias & Type I error**: `vignette("bias-and-type1")`
- **Time-to-event endpoints**: `vignette("tte-endpoints")`
