---
title: "Time-to-Event Endpoints and Surrogacy Framework"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time-to-Event Endpoints and Surrogacy Framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(dosopt)
```

## Overview

Dose optimization studies often use a short-term binary endpoint (e.g., response rate) for Stage 1 selection while the confirmatory Stage 2 trial uses a time-to-event (TTE) endpoint (e.g., progression-free survival, overall survival). This vignette covers how **dosopt** quantifies selection bias for TTE endpoints via a surrogacy framework.

## Surrogacy Framework

Let $X$ be the binary surrogate (efficacy rate) used for dose selection, and $T$ be the TTE endpoint used in the confirmatory trial. The surrogacy correlation $\rho_{TX}$ links the bias in $X$ to bias in $T$:

$$|\text{Bias}(\bar{T})| \leq |\rho_{TX}| \cdot \frac{\sigma_T}{\sigma_X} \cdot |\text{Bias}(\hat{p})|$$

Setting $\rho_{TX} = 1$ gives a conservative upper bound.

## TTE Bias Bounds

```{r tte-bias}
# Stage 1 binary bias (from sample size design)
ss <- calc_sample_size_utility_approx(
  pL = 0.30, qL = 0.70, delta = 0.10, d = 0.15,
  phi = 0, alpha_L = 0.80, alpha_H = 0.80
)
bin_bias <- calc_bias(
  pL = 0.30, qL = 0.70, delta = 0.10, d = 0.15,
  phi = 0, n1 = ss$n, n2 = 2 * ss$n, lambda_u = ss$lambda_u
)

# TTE bias bounds assuming exponential survival, median = 12 months
tte_bias <- calc_bias_TTE(
  bias_binary  = bin_bias$bias_stage1,
  p0           = 0.30,           # surrogate response rate under H0
  lambda0      = log(2) / 12,    # TTE hazard under H0
  rho_TX       = 0.7,            # surrogate-TTE correlation
  n1           = ss$n
)
print(tte_bias)
```

The function returns both the surrogate-adjusted estimate (`bias_TTE`) and the conservative upper bound (`bias_TTE_upper`).

## Expected Events Under Exponential + Uniform Accrual

For trial planning purposes, compute expected events under exponential survival with uniform accrual:

```{r expected-events}
ev <- calc_expected_events(
  n        = 200,           # total patients
  lambda   = log(2) / 12,  # hazard (median 12 months)
  t_accrual = 18,           # accrual period (months)
  t_follow  = 12            # minimum follow-up (months)
)
print(ev)
```

## Exponential Log-Scale Type I Error

For a one-arm test comparing estimated hazard $\hat{\lambda}$ to a null $\lambda_0$ on the log scale:

```{r t1e-exp}
t1e_exp <- calc_type1_error_exp(
  bias_stage1  = bin_bias$bias_stage1,
  n1           = ss$n, n2 = 2 * ss$n,
  p0           = 0.30,
  rho          = 0.7,
  lambda0      = log(2) / 12,
  n_events     = 80,
  alpha        = 0.05
)
print(t1e_exp)
```

## Two-Arm Cox Regression

The most common confirmatory design uses a Cox proportional hazards model. Bias propagates through a delta-method chain:

$$\text{Bias}(\hat{p}) \xrightarrow{\text{surrogacy}} \text{Bias}(\hat{\lambda}) \xrightarrow{\log} \text{Bias}(\log \widehat{HR}) \xrightarrow{\text{SE}} \text{Bias}(Z_{\text{Cox}})$$

```{r t1e-cox}
t1e_cox <- calc_type1_error_cox(
  bias_stage1 = bin_bias$bias_stage1,
  n1          = ss$n, n2 = 2 * ss$n,
  p0          = 0.30, rho = 0.7,
  lambda0     = log(2) / 12,
  n_confirm   = 300,
  alpha       = 0.05
)
print(t1e_cox)
```

## Monte Carlo Validation with TTE Endpoints

The `simulate_dose_opt()` function supports TTE endpoints with Gaussian copula dependence between the binary surrogate and the TTE outcome:

```{r simulation, eval = FALSE}
sim_res <- simulate_dose_opt(
  pL = 0.30, qL = 0.70, delta = 0.10, d = 0.15,
  phi = 0, n1 = ss$n, n2 = 2 * ss$n,
  endpoint = "TTE",
  lambda0  = log(2) / 12,
  rho_TX   = 0.7,
  n_sim    = 1000,
  seed     = 42
)
print(sim_res)
```

The simulation returns empirical bias and Type I error alongside the analytical approximations, allowing cross-validation of the closed-form results.

## Practical Recommendations

When the confirmatory endpoint is TTE:

1. Use `calc_bias_TTE()` with a range of `rho_TX` values (e.g., 0.5–0.9) to assess sensitivity.
2. Report both the surrogate-adjusted estimate and the upper bound ($\rho_{TX} = 1$) in study protocols.
3. If analytical Type I error exceeds 0.10, consider adjusting the nominal confirmatory alpha or increasing $n_1$.
